Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
rule permutation

Fix pairwise disjoint countably infinite sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

\subsection{Rule permutation for the proof of Lemma~\ref{lem:cw-decomposition}}
\label{app:cw-decomposition}

We first study the interactions between two rules (only non-trivial cases are
presented here):
\begin{itemize}
\item $\rr_1/\rr_2$, where $\rr_1 \in \set{\wrD, \wfaD}$ and $\rr_2 \in
\set{\acD, \cfaD, \me, \mfaD, \mexD}$:

@@

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.A}{
    \vlin{\wfaD}{(x \notin fv(\forall x.A))}{\forall x.\forall x.A}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlhy{\forall x.A}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{(x \notin fv(A))}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{A \vlor (\forall x.B)}}}}
\leadsto
\vlderivation{
  \vlin{\fequ}{(x \notin fv(A))}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor (\forall x.B)}}} 
\end{equation*}

@@

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{(\forall x.A) \vlor B}{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlhy{\forall x.\forall x.(A \vlor B)}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
    \vlin{\fequ}{(x \notin fv(B))}{(\forall x.\forall x. A) \vlor B}{
      \vlhy{\forall x.\forall x.(A \vlor B)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{\forall x.(A \vlor B)}{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlhy{(\forall x.\forall x.A) \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{(x \notin fv(B))}{\forall x.\forall x.(A \vlor B)}{
      \vlhy{(\forall x.\forall x.A) \vlor B}}}}
\end{equation*}

@@

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{(\forall x.A) \vlor B}{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{(\forall x. A) \vlor B}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{(\forall x.A) \vlor B}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\forall x.(A \vlor B)}{
    \vlhy{\forall x.A}}}
\end{equation*}

\item $\wfaD/\fequ$:

@@

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{(\forall x.A) \vlor B}{
    \vlin{\wfaD}{(x \notin fv(A \vlor B))}{\forall x.(A \vlor B)}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(A))}{(\forall x.A) \vlor B}{
    \vlhy{A \vlor B}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{(x \notin fv(B))}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{(x \notin fv(A))}{(\forall x.A) \vlor B}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin fv(A \vlor B))}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor B}}}
\end{equation*}

@@

\item $\wrD/\fequ/\cfaD$: We only list non-trivial cases here:
  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlin{\fequ}{(x \notin fv(B))}{(\forall x.\forall x.A) \vlor B}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{(\forall x.A) \vlor B}{
      \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlin{\fequ}{(x \notin fv(B))}{(\forall x.\forall x.A) \vlor B}{
        \vlin{\wrD}{}{\forall x.(B \vlor (\forall x.A))}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{(x \notin fv(B))}{(\forall x.A) \vlor B}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{(x \notin fv(B))}{\forall x.\forall x.(A \vlor B)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlin{\cfaD}{}{\forall x.A}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{(x \notin fv(B))}{\forall x.\forall x.(A \vlor B)}{
        \vlin{\wrD}{}{\forall x.(B \vlor (\forall x.A))}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.(A \vlor B)}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}

\item $\wrD/\fequ/\acD$:
  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{a \vlor B}{
      \vlin{\fequ}{}{(a \vlor a) \vlor B}{
        \vlin{\wrD}{}{(a \vlor B) \vlor a}{
          \vlhy{a \vlor B}}}}}
  \leadsto
  \vlderivation{
    \vlhy{a \vlor B}}
  \end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\fequ}{(x \notin fv(B \vlan D))}{\forall x.((A \vlan C) \vlor (B \vlan
D))}{
        \vlin{\wrD}{}{(B \vlan D) \vlor (\forall x.(A \vlan C))}{
          \vlhy{B \vlan D}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan (D \vlor C))}{
        \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan D)}{
          \vlin{\faD}{(x \notin fv(B \vlan D))}{\forall x.(B \vlan D)}{
            \vlhy{B \vlan D}}}}}}
  \end{equation*}

@@

>>>>

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

\subsection{Rule permutation for the proof of Lemma~\ref{lem:cw-decomposition}}
\label{app:cw-decomposition}
We construct a rewriting system based on rule permutation on derivations in $\set{\wrD, \wfaD, \acD,
\cfaD, \me, \mfaD, \mexD, \fequ}$ that allows us to reach a derivation of the
form 
\begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{B}{
        \vlde{\set{\acD,\cfaD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
\end{equation*}
from any derivation. Intuitively, we want to move all the instances of $\rr \in
\set{\wrD, \wfaD}$ downwards and all the instances of $\rr' \in \set{\me,
\mfaD, \mexD}$ upwards.

We first study the interactions between two rules. Certain cases are unsolved at
this stage, and they are considered later when we study the interactions between
two non-$\fequ$ rule instances seperated by $\fequ$. Only non-trivial cases are
considered here:
\begin{itemize}
\item $\rr_1/\rr_2$, where $\rr_1 \in \set{\wrD, \wfaD}$ and $\rr_2 \in
\set{\acD, \cfaD, \me, \mfaD, \mexD}$:

@@

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.A}{
    \vlin{\wfaD}{}{\forall x.\forall x.A}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlhy{\forall x.A}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{A \vlor (\forall x.B)}}}}
\leadsto
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor (\forall x.B)}}} 
\end{equation*}
where in the last case, $x$ is not free in $A$.

@@

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlhy{\forall x.\forall x.(A \vlor B)}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
    \vlin{\fequ}{}{(\forall x.\forall x. A) \vlor B}{
      \vlhy{\forall x.\forall x.(A \vlor B)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlhy{(\forall x.\forall x.A) \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{}{\forall x.\forall x.(A \vlor B)}{
      \vlhy{(\forall x.\forall x.A) \vlor B}}}}
\end{equation*}
where in the last two cases, $x$ is not free in $B$.

@@

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{(\forall x. A) \vlor B}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\wrD}{}{\forall x.(B \vlor A)}{
      \vlhy{\forall x.B}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{(\forall x.A) \vlor B}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\forall x.(A \vlor B)}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{B \vlor (\forall x.A)}{
      \vlhy{B}}}}
\end{equation*}
where in the last four cases, $x$ is not free in $B$.
\item $\wfaD/\fequ$:

@@

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\wfaD}{}{\forall x.(A \vlor B)}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{}{(\forall x.A) \vlor B}{
    \vlhy{A \vlor B}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{}{(\forall x.A) \vlor B}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor B}}}
\end{equation*}
where in the last two cases, the constraint on $x$ on the left-hand side implies
that of the right-hand side.

@@

\item $\wrD/\fequ/\cfaD$:
  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlin{\fequ}{}{(\forall x.\forall x.A) \vlor B}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{(\forall x.A) \vlor B}{
      \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlin{\fequ}{}{(\forall x.\forall x.A) \vlor B}{
        \vlin{\wrD}{}{\forall x.(B \vlor (\forall x.A))}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{(\forall x.A) \vlor B}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{\forall x.\forall x.(A \vlor B)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlin{\cfaD}{}{\forall x.A}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{\forall x.\forall x.(A \vlor B)}{
        \vlin{\wrD}{}{\forall x.(B \vlor (\forall x.A))}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.(A \vlor B)}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}
  where in all four cases, $x$ is not free in $B$.
\item $\wrD/\fequ/\acD$:
  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{a \vlor B}{
      \vlin{\fequ}{}{(a \vlor a) \vlor B}{
        \vlin{\wrD}{}{(a \vlor B) \vlor a}{
          \vlhy{a \vlor B}}}}}
  \leadsto
  \vlderivation{
    \vlhy{a \vlor B}}
  \end{equation*}

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\fequ}{}{\forall x.((A \vlan C) \vlor (B \vlan
D))}{
        \vlin{\wrD}{}{(B \vlan D) \vlor (\forall x.(A \vlan C))}{
          \vlhy{B \vlan D}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan (D \vlor C))}{
        \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan D)}{
          \vlin{\faD}{}{\forall x.(B \vlan D)}{
            \vlhy{B \vlan D}}}}}}
  \end{equation*}
  where in the second case, $x$ is free in $B \vlan D$.

@@

  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.A) \vlor (\forall x.B)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlin{\cfaD}{}{\forall x.A}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}
